---
title: "Class 10: Mini Project"
author: "Zainub Darsot: A16294217"
format: pdf
---

Today we will be looking at some data regarding Candy Stats. 


```{r}
#read candy-data file  

candy_file <- "candy-data.csv"

#omit the first row 
candy = read.csv(candy_file, row.names=1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
#use `nrow` to count the rows of candy 
nrow(candy)
```

## Q2. How many fruity candy types are in the dataset?
```{r}
#use sum to count the number of candy types (columns) 

sum(candy$fruity)
```

## Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Almond Joy", ]$winpercent
```

## Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Almond Joy", ]$winpercent
```


## Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```

```{r}
#load skimr 
library("skimr")
skim(candy)
```


## Question Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent seems to be on a different scale to the majority of the other columns in the dataset. 

## Question Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy[as.logical(candy$chocolate), ]
```
1 means that there is chocolate in the candy and 0 means there is no chocolate. 

## Question Q8. Plot a histogram of winpercent values
```{r}
#load ggplot 
library(ggplot2)

```

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=10)
```


## Question Q9. Is the distribution of winpercent values symmetrical?
The distribution seems to be right skewed, not symmetrical. 

## Question Q10. Is the center of the distribution above or below 50%?
The center of distribution is below 50%.

## Question Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
choc_candy <- candy[which(candy$chocolate == TRUE), ]
fruit_candy <- candy[which(candy$fruity == TRUE), ]

mean(choc_candy$winpercent) > mean(fruit_candy$winpercent)

```
OR 

```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds, "winpercent"]

fruit.inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.inds, "winpercent"]

mean(choc.win) > mean(fruit.win)

```

On average, chocolate candy is higher ranked than fruity candy. 

## Question Q12. Is this difference statistically significant?
```{r}
#use t.test to find significance 
t.test(choc_candy$winpercent, fruit_candy$winpercent )
```
The difference is statistically significant. 

## Question 13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```
The 5 least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

## Question 14. What are the top 5 all time favorite candy types out of this set?
```{r}
#load dplyr
library("dplyr")
tail(candy[order(candy$winpercent),], n=5)
```
The top 5 are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter Cups

## Question 15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy), winpercent) +
  geom_col(fill="grey")
```

## Question 16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill="grey")
```

```{r}
#rename so that we can see colors 
mycols <- rep("grey", nrow(candy))

mycols[as.logical(candy$chocolate)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
```

```{r}
#generate plot
ggplot(candy) + 
  aes(winpercent, rownames(candy), winpercent) +
  geom_col(fill=mycols)
```


Now, for the first time, using this plot we can answer questions like:
## Question 17. What is the worst ranked chocolate candy?
The worst ranked chocolate candy is Sixlets. 

## Quesiton 18. What is the best ranked fruity candy?
The best ranked fruit candy is Starburst.

```{r}
#install.packages("ggrepel")
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 7)
```

## Question 19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
#using order, find the price and most ranked 
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12)], n=5 )
```
The most liked and least expensive is Tootsie Roll Midgies. 

## Question 20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
#
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The most expensive candies are Nik L Nip, Nestle Smarties, Ring Pop, Hersheys Krackel, and Hersheys Milk Chocolate. Of these, Hersheys Milk Chocolate is the most expensive.

```{r}
#install.packages("corrplot")

#load corrplot 
library(corrplot)

```

```{r}
cij <- cor(candy)
corrplot(cij)
```

## Question 22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Based on the plot, the two most anti-correlated variables are Fruit and Chocolate. 


## Question 23. Similarly, what two variables are most positively correlated?
Based on the plot above, the two variables most positvely correlated are chocolate and winpercent. 


```{r}
#generate pca
pca <- prcomp(candy, scale=T)
summary(pca)
```


```{r}
#plot PC1 vs PC2

plot(pca$x[,1:2])

```

```{r}
#add color
plot(pca$x[,1:2], col=mycols, pch=16)

```

```{r}
#load ggplot2
library(ggplot2)
my_data <- cbind(candy, pca$x[,1:3])
my_data
```


```{r}
#generate plot 
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```



```{r}
library("plotly")
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

## Question 24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Fruity, Hard and Pluribus variables are picked up strongly by PC1 in the positive direction. This makes sense because they are positively correlated. 


