---
title: "Class 10.1: Comparative structure analysis of Adenylate Kinase"
author: "ZainubDarsot (PID:A16294217)"
format: pdf
---

We need some packages. These include `bio3d` and `msa`. 

The `msa` package is from BioConductor. These pacakages focus on genomics type work and are managed by the `Biocmanager package`

Install `install.packages("BiocManager")`
`BiocManager:: install("msa")` all entered in the R "brain" console.

```{r}
#load bio3d

library(bio3d)

#retrieve amino acid sequence
aa <- get.seq("1ake_A")

aa
```

Now I can search the PDB database for related sequences:  

```{r}
# Blast or hmmer search 
#b <- blast.pdb(aa)
```

```{r}
# Plot a summary of search results
#hits <- plot(b)
```


```{r}
#attributes(b)
#head(b$hit.tbl)
```

These are the related structures in the PDB database that we found via a BLAST search...

Load top-scoring hits: 
```{r}
hits <- NULL
hits$pdb.id <- c("1AKE_A","8BQF_A", "4X8M_A", "6S36_A", "6RZE_A", "4X8H_A", "3HPR_A", "1E4V_A", "5EJE_A", "1E4Y_A", "3X2S_A", "6HAP_A", "6HAM_A", "4K46_A", "4NP6_A", "3GMT_A","4PZL_A")
```

Side-note: Lets annotate these structures (aka: find out what they are, what species they are from, stuff about the experiment they were solved in etc.)

Use `pdb.annotate()` function: 
```{r}
annotate.hits <- pdb.annotate(hits$pdb.id)
annotate.hits
```


```{r}
attributes(annotate.hits)
```

```{r}
annotate.hits <- pdb.annotate(hits$pdb.id)
```

```{r}
#attributes(annotate.hits)
head(annotate.hits)
```

Now we can download all these structures for further analysis with the `pdb.get()` function. 
```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Now we have all the PDB related structures we can Align and Supperpose  ...

```{r}
#Align related PDB's
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
pdbs
```

## Principle Component Analysis: 

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)

```

```{r}
# Visualize PC1
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
pc1
```